---
title: 'Structural Equation Modeling: Holzinger and Swineford Dataset'
author: "Felipe Villota"
date: "2/2/2022"
output:
  word_document: default
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r include=FALSE}
 # LAB ASSIGNMENT 4 - STRUCTURE EQUATION MODELLING:  MENTAL ABILITY TEST SCORES
# Luis Felipe Villota Macías 

# Objective: to evaluate fit of a theoretical factor structure of mental ability 
# test indicators and relationships between different test scores. It is a
# theory-based confirmatory research. 



#Loading packages
library(lavaan) # for SEM fit and model functions	
library(semPlot) # for semPaths()	
library(semptools) # for set_sem_layout	
library(tidyverse) # for tidy code	
library(CompQuadForm) # for mvnorm.kur.test and mvnorm.skew.test (prerequisite)	
library(ICS) # for mvnorm.kur.test and mvnorm.skew.test	
library(psychTools)
library(psych)	
library(readr)
library(skimr)
library(Amelia)
library(MVN)
library(semTable)

# Setting working directory
getwd()
setwd("C:/Users/USER/Desktop/SSDA - LUND UNIVERSITY/- COURSES/SIMM61- Quantitative Data Analysis in R/Scripts/Zoltan")

# Loading the data 

mental_data = holzinger.swineford

# Checking the data
View(mental_data)
summary(mental_data)
skim(mental_data)
describe(mental_data) # 301 obs 33 vars

# Missing values
any(is.na(mental_data)) # TRUE
sum(is.na(mental_data)) # 312 missing cases in total
colSums(is.na(mental_data)) # missing values in "t25_frmbord2" and "t26_flags"
missmap(mental_data)
# We won't remove any NAs, it will reduce the dataset considerably.

# Variables of interest: tidying and recoding 
mental_data = mental_data %>% 
                     rename(sex = female) # female to sex

# _________________________

# Path diagrams of the MEASUREMENT MODEL: the latent variable (here exogenous)
# is the unobserved cause of manifest variables (here endogenous). It is a 
# latent variable loading in SEM.

# From lab assignment: "Your theoretical model indicates that there are three (3) 
# underlying factors that influence the mental ability test scores.These latent 
# factors can correlate with each other.

# 1) Visual perception ability is a latent factor that influences (causally determines) test scores 
# measured by t01_visperc, t02_cubes, t03_frmbord, and t04_lozenges, which are all related to 
# visual perception and mental rotation.


model_visper <- '
visper =~ t01_visperc + t02_cubes + t03_frmbord + t04_lozenges
'
fit_visper <- sem(model_visper, data = mental_data)
plot = semPaths(fit_visper, label.scale=F, nCharNodes = 8,
                sizeMan2=3.5, sizeMan=10, asize=3, edge.color="black", residuals = F, fixedStyle = 1)


# 2) Verbal ability is a latent factor that influences (causally determines) test scores measured 
# by t06_paracomp, t07_sentcomp, and t09_wordmean.

model_verbal <- '
verbal =~ t06_paracomp + t07_sentcomp + t09_wordmean 
'
fit_verbal <- sem(model_verbal, data = mental_data)
plot = semPaths(fit_verbal, label.scale=F, nCharNodes = 8,
                sizeMan2=3.5, sizeMan=10, asize=3, edge.color="black", residuals = F, fixedStyle = 1)



# 3) Processing speed is a latent factor that influences (causally determines) test scores 
# measured by t10_addition t12_countdot, and t13_sccaps." 

model_processing <- '
processing =~ t10_addition + t12_countdot + t13_sccaps 
'
fit_processing <- sem(model_processing, data = mental_data)
plot = semPaths(fit_processing, label.scale=F, nCharNodes = 8,
                sizeMan2=3.5, sizeMan=10, asize=3, edge.color="black", residuals = F, fixedStyle = 1)

#______________________________



# MODEL A

modelA<- '
        # measurement model
          visper =~ t01_visperc + t02_cubes + t03_frmbord + t04_lozenges
          verbal =~ t06_paracomp + t07_sentcomp + t09_wordmean 
          processing =~ t10_addition + t12_countdot + t13_sccaps
'

fit_modelA <- sem(modelA, data = mental_data)
fit_modelA

# unstandardized

summary(fit_modelA)

plot = semPaths(fit_modelA, label.scale=F, nCharNodes = 8,
                sizeMan2=3.5, sizeMan=10, asize=3, edge.color="black", residuals = F, fixedStyle = 1,
                whatLabels = "est")

semPaths(fit_modelA, whatLabels = "est")

# standardized 

summary(fit_modelA, standardized = T, rsquare = T)

plot = semPaths(fit_modelA, label.scale=F, nCharNodes = 8,
                sizeMan2=3.5, sizeMan=10, asize=3, edge.color="black", residuals = F, fixedStyle = 1,
                whatLabels = "std")

semPaths(fit_modelA, whatLabels = "std")

#______________________________

# Counting degrees of freedom: 

summary(fit_modelA) # df = 32, Model A is identified 

# From the notes: "difference between the total number of possible estimated 
# parameters of a “null model” with no latent variables, and the number of free parameters
# in our actual model". 

# 10 manifest variables
# p*(p+1)/2

10*(10+1)/2 # total possible parameters of the manifest variables = 55 



parTable(fit_modelA)
parameterEstimates(fit_modelA)

# The number of free parameters: 

# 10 latent factor loadings between:
# visper =~ t01_visperc + t02_cubes + t03_frmbord + t04_lozenges
# verbal =~ t06_paracomp + t07_sentcomp + t09_wordmean 
# processing =~ t10_addition + t12_countdot + t13_sccaps
# 3 covariances (between our exogenous variables)
# 3 variances (for each latent factor)
# 7 residuals

# 
# That gives us a df = 55 - 23 = 32   

#______________________________

# Checking assumptions: MULTIVARIATE NORMALITY 


## Correlations: between all items in the questionnaire

mental_tscores = mental_data %>% dplyr::select(t01_visperc:t26_flags)  #database with 26 test scores
cor_mental_tscores <- mental_tscores %>% # correlations between items
        cor()
cor_mental_tscores




# Multivariate normality through the Henze-Zirkler's test

mental_mvn <- mvn(mental_data[,c("t01_visperc", "t02_cubes", "t03_frmbord", "t04_lozenges", 
                                 "t06_paracomp", "t07_sentcomp", "t09_wordmean", 
                                 "t10_addition", "t12_countdot", "t13_sccaps")], mvnTest = "hz") 
mental_mvn$multivariateNormality   # p <0.05, MVN is violated


# Multivariate Normality Test Based on Kurtosis
mvnorm.kur.test(mental_data[,c("t01_visperc", "t02_cubes", "t03_frmbord", "t04_lozenges", 
                           "t06_paracomp", "t07_sentcomp", "t09_wordmean", 
                           "t10_addition", "t12_countdot", "t13_sccaps")])  # p<0.05, MVN is violated

# Multivariate Normality Test Based on Skewness
mvnorm.skew.test(mental_data[,c("t01_visperc", "t02_cubes", "t03_frmbord", "t04_lozenges", 
                               "t06_paracomp", "t07_sentcomp", "t09_wordmean", 
                               "t10_addition", "t12_countdot", "t13_sccaps")])  # p<0.05, MVN is violated

# All MVN tests show p values lower than 0.05, so the assumption of MVN is violated. 
# P values are not reliable, the standard errors are large, model fit indices are unreliable.
# We proceed with the robust variants of the statistics of the model fit: e.g Satorra-Bentler 
# (corrects Chi-squared fit), normality-adjusted robust standard errors. 


# ML with robust SE and test statistics
fit_modelA_MLM <- sem(modelA, data = mental_data, estimator = "MLM")
summary(fit_modelA_MLM, fit.measures = T)

# Bootstrapped ML with robust SE and test statistics
fit_modelA_BOOT <- sem(modelA, data = mental_data, se = "bootstrap", test = "bootstrap")
summary(fit_modelA_BOOT)





#_______________________________________


# MODEL B 

modelB<- '
        # measurement model
          visper =~ t01_visperc + t02_cubes + t03_frmbord + t04_lozenges
          verbal =~ t06_paracomp + t07_sentcomp + t09_wordmean 
          processing =~ t10_addition + t12_countdot + t13_sccaps
          t10_addition~~t12_countdot
'



fit_modelB_MLM <- sem(modelB, data = mental_data, estimator = "MLM")
fit_modelB_MLM


# unstandardized

summary(fit_modelB_MLM)

plot = semPaths(fit_modelB_MLM, label.scale=F, nCharNodes = 8,
                sizeMan2=3.5, sizeMan=10, asize=3, edge.color="black", residuals = F, fixedStyle = 1,
                whatLabels = "est")

semPaths(fit_modelB_MLM, whatLabels = "est")

# standardized 

summary(fit_modelB_MLM, standardized = T, rsquare = T)

plot = semPaths(fit_modelB_MLM, label.scale=F, nCharNodes = 8,
                sizeMan2=3.5, sizeMan=10, asize=3, edge.color="black", residuals = F, fixedStyle = 1,
                whatLabels = "std")

semPaths(fit_modelB_MLM, whatLabels = "std")


standardizedsolution(fit_modelB_MLM)


# Indices of Model B
summary(fit_modelB_MLM, fit.measures = T)


#_______________________________________________________


# Reproduction of Task 3 Path Diagram 


modelC<- '
        # measurement model
          t13_sccaps ~ c*t01_visperc + b*t12_countdot 
          t12_countdot ~ a*t01_visperc 
          
        # indirect effect (a*b)
            indirect := a*b
        
        # total effect
            total := c + (a*b)
          
'


fit_modelC <- sem(modelC, data = mental_data)

summary(fit_modelC)

plot = semPaths(fit_modelC, label.scale=F, nCharNodes = 8,
                sizeMan2=3.5, sizeMan=15, asize=3, edge.color="black", residuals = F, fixedStyle = 1,
                edge.label.cex = 1)



fit_modelC_MLM <- sem(modelC, data = mental_data, estimator = "MLM")
fit_modelC_MLM

semPaths(fit_modelC_MLM, whatLabels = "est")

summary(fit_modelC_MLM, fit.measures = T)

parameterestimates(fit_modelC)

```

## Introduction

Our objective in the present report is to to evaluate the fit of a theoretical factor structure of intelligence-test indicators and their relationships between different test scores based on the Holzinger and Swineford [dataset](https://www.rdocumentation.org/packages/psychTools/versions/2.0.8/topics/holzinger.swineford) through the structural equation modeling statistical technique. It is a theory-based confirmatory research which, in this exercise, covers (3) established latent and exogenous factors of *visual perception ability*, *processing speed*, and *verbal ability* underlying and causally determining the scores of participants in 10 different tests (endogenous and manifest variables) (below described).

## Data and Methods

***Data cleaning***. In the full database we have 301 observations and 33 variables. We had 312 missing values in total in variables "*t25_frmbord2*" and "*t26_flags*". We didn't remove any *NA*s, because it would've reduced the observations considerably and those were not variables of interest here. We renamed variable "*female*" into "*sex*".

***Model A specification.*** We built Model A according to the theoretical assumptions in which the measurement model was composed of three (3) exogenous variables (here latent factors) and 10 endogenous variables (manifest variables) in the following manner: a.) *Visual perception* causally determines test scores in "*t01_visperc*", "*t02_cubes*", "*t03_frmbord*", and "*t04_lozenges*"; b.) *Verbal ability* causally determines test scores in "*t06_paracomp*", "*t07_sentcomp*", and "*t09_wordmean*"; c.) *Processing speed* causally determines test scores in "*t10_addition*", "*t12_countdot*" and "*t13_sccaps"*. Finally, these three exogenous variables ( latent factors) are correlated to one another.

***Checking assumptions.*** All multivariate normality tests (*Henze-Zirkler*, *MVN based on kurtosis*, *MVN based on skewness*) show significant p values (\< 0.05), so the assumption of MVN is violated. For our model, it means that the model fit indices are not reliable.

In this line, we used the Maximum likelihood estimator, with robust SE and test statistics to find the most likely solution for Model A. Although it assumes MVN, it allows unstandardized observed variables, it is scale-free (the test scores have different scales).

We proceeded with the normality-adjusted robust variants of the statistics: The model Chi-squared = 0.00, Satorra-Bentler (corrects Chi-squared fit) = 1.047, degrees of freedom = 32, p-value = 0.00, TLI = 0.917, CFI = 0.941 and RMSEA =0.075, with 90 Percent confidence interval in the lower = 0.056, and the upper= 0.095. We find the model is acceptable: is identified df \>1; and according to the predominant literature of Hu and Bentler's (1999), the model shows acceptable goodness of fit indices: TLI \< 0.95, CFI \< 0.95, SRMR = 0.062 \< 0.08.

***Model B specification.***We include in our previous model a suspected correlation between "*t10_addition*" and "*t12_countdot*" (not determined by *processing speed*). This was due to the realization that these two variables had a similar structure (distinct from the one of "*t13_sccaps*") so it was deduced that after accounting for the influence of *processing speed* some possible correlation would remain between the mentioned two variables above.

## Results

***Comparison of Model A and Model B.*** Since there is no previous findings to support theoretically which model (A or B ) is the best, we compared them in order to see which fits better the data: Model A ( AIC= 8296.856, BIC= 8382.120, SABIC= 8309.177) and Model B ( AIC= 8267.652, BIC= 8356.623, SABIC = 8280.508). There is a significant difference between the model fits, the AIC of both have a difference \>2. The accepted criteria regarding this indicator tells us to prefer the model with the smallest AIC, which in this case is Model B. Regarding BIC and SABIC, and following a similar logic of comparison, we have again Model B as a better fit for the data. \
\
***Influence of the Visual Perception Ability factor in Model B.*** For this final model, we interpret the standardized solution estimates. We found that "*t02_cubes" (Cubes, Simplification of Brighams Spatial Relations Test score)* is the manifest variable that was the least influenced by the visual perception ability factor (latent variable). This is based on its factor loading = 0.442 and its R\^2 = 0.195 (\~20% of variance of the visual perception ability factor explained).

In Figure i, we observe the full path diagram of Model B with standardized estimates.The rest of the manifest variables factor loadings are: "*t01_visperc*" = 0.751, "*t04_lozenges*" = 0.596 and "*t03_frmbord" =* 0.495.

Figure i. Plot of Path Diagram of Model B (with standardized estimates)

```{r}

semPaths(fit_modelB_MLM, whatLabels = "std")

```

## Discussion of Task 3 Path Diagram

We reproduced the path diagram proposed through the creation of Model C, in which "*t13_sccaps*" is a manifest and endogenous variable, determined by the endogneous variable "*t01_visperc*" mediated by "*t12_countdot*". Using the "MLM" estimator, we have 0 degrees of freedom and the Test statistic =0.000.

If we look to the unstandardized estimates of the modeled paths, we see that if "*t01_visperc*" is increased by 1 unit, it is foreseen a total change of 0.394 in "*t13_sccaps*". The reasoning behind this claim is based on the model relationships, in which "*t01_visperc"* has a direct effect (labeled "*c" =* 0.308) on the dependent variable. It has also indirect effect (labeled and in the form of "*a\*b" =* 0.379 *\** 0.227) on the outcome mediated by "*t12_countdot*". So, 1-unit increase of "*t01_visperc*" would suppose a total influence *c+(a\*b)*.

Figure ii. Plot of Path Diagram of Model C (with unstandardized estimates)

```{r}
semPaths(fit_modelC_MLM, whatLabels = "est")

```

\
Figure ii. Plot of Path Diagram of Model C (with mediation names)

```{r}
plot = semPaths(fit_modelC, label.scale=F, nCharNodes = 8,
                sizeMan2=3.5, sizeMan=15, asize=3, edge.color="black", residuals = F, fixedStyle = 1,
                edge.label.cex = 1)
```
